NOPT048 Notes

### 2019-09-15 misc/study-notes/groups.tex
Edited group theory notes

% Abstract Algebra Notes
% Based on lectures given in Fall 2003 by Prof. Benedict Gross
% Notes by Marcel Goh

@@

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}

\leftrighttop{NOTES ON ABSTRACT ALGEBRA}{MARCEL GOH}

\maketitlenodate{Notes on Abstract Algebra\footnote{$^1$}{\eightpt Based on a series of lectures given at Harvard University in Fall 2003 by Prof. Benedict Gross.}}{by}{Marcel K.\ Goh}

@@

Thus the set $GL_n(\R)$, under the operation of matrix multiplication, has a multiplicative inverse $A^{_1}$ for every matrix $A$. It also contains an identity element $I$ and the multiplication operation is associative. These are the properties of a group.

@@

A {\it group} $G$ is a set on which is defined a rule of combination such that the product of two elements $g,h\in G$, denoted $g\cdot h$ or $gh$, is also in $G$. Furthermore, the following three properties must hold:
\medskip
\item {a)} Multiplication must be associative: for all $g,h,k\in G$, $(gh)k = g(hk)$.
\smallskip
\item {b)} There exists an identity element $e\in G$ such that $ge = eg = g$ for all $g\in G$. This element is also often denoted $1$.
\smallskip
\item {c)} For every element $g\in G$, there exists an inverse element $g^{-1}$ such that $gg^{-1} = g^{-1}g = e$.
\medskip
If $gh = hg$ for all $g,h\in G$, then we say the group $G$ is {\it commutative} or {\it abelian}. The number of elements in $G$ is called its {\it order} and this value, also denoted $|G|$, need not be finite.

The most familiar abelian group is the set of all integers, denoted $\Z$, under the operation of addition. It has $1$ as its identity, inverses $-a$ for every whole number $a$, and the commutative property; likewise, any vector space $V$ is also an abelian group under vector addition.

Group theory is intimately connected to the study of symmetries of an object. Let $T$ be a set and let $\Sym(T)$ denote the set of all bijections from $T$ to itself. This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $e$ and every bijection $f$ has an inverse bijection $f^{-1}$. In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections. For instance, $GL_n(\R) \subset \Sym(\R^n)$. If $T = \{1,\ldots,n\}$, then $\Sym(T)$ is called the symmetric group on $n$ letters and is denoted $S_n$.

\section 2.2 Subgroups

@@

We turn our attention to $S_3$, the permutation group on three letters. Notice that this group is not commutative, since the elements
$$\sigma = \pmatrix{1 & 2 & 3\cr 2 & 3 & 1} \quad \hbox{and} \quad
  \tau = \pmatrix{1 & 2 & 3\cr 2 & 1 & 3}$$
do not commute:
$$\sigma\tau = \pmatrix{1 & 2 & 3 \cr 3 & 2 & 1}, \quad \hbox{but} \quad
  \tau\sigma = \pmatrix{1 & 2 & 3 \cr 1 & 3 & 2}.$$
Note that for $k\leq n$, $S_k\leq S_n$ because we can simply fix the letters $k+1,k+2,\ldots,n$. An easy corollary, then, is that $S_n$ does not commute for $n\geq 3$. This is because $S_3\leq S_n$ and we can simply take $\sigma$ and $\tau$ as elements of the larger group $S_n$ that do not commute.

@@

\section 2.3 Isomorphisms and Homomorphisms

@@

The {\it kernel} of a homomorphism is a the set of all elements in the domain that map to the identity in the target group. The {\it image} of a homomorphism $f: G_1\rightarrow G_2$ is the set of all elements in $G_2$ that equal $f(x)$ for some $x\in G_1$.

@@

>>>>

% Group Theory Notes
% Notes by Marcel Goh

@@

\def\Sym{\mathop{\rm Sym}\nolimits}
\def\Aut{\mathop{\rm Aut}\nolimits}
\def\Im{\mathop{\rm Im}\nolimits}
\def\Id{\mathop{\rm Id}\nolimits}

\leftrighttop{NOTES ON THE THEORY OF GROUPS}{MARCEL GOH}

\maketitlenodate{Notes on the Theory of Groups}{by}{Marcel K.\ Goh}

@@

Thus the set $GL_n(\R)$, under the operation of matrix multiplication, has a multiplicative inverse $A^{-1}$ for every matrix $A$. It also contains an identity element $I$ and the multiplication operation is associative. These are the properties of a group.

@@

A {\it group} $G$ is a set on which is defined a rule of combination such that the product of two elements $g,h\in G$, denoted $g\cdot h$ or $gh$, is also in $G$. Furthermore, the following three properties must hold:
\medskip
\item {a)} Multiplication must be associative: for all $g,h,k\in G$, $(gh)k = g(hk)$.
\smallskip
\item {b)} There exists an identity element $e\in G$ such that $ge = eg = g$ for all $g\in G$. This element is also often denoted $1$.
\smallskip
\item {c)} For every element $g\in G$, there exists an inverse element $g^{-1}$ such that $gg^{-1} = g^{-1}g = e$.
\medskip
An immediate consequence of the axioms is that the identity $e$ is unique. For if both $e$ and $e'$ are the identity of a group, then $e = ee' = e'$.

The term {\it order} serves a somewhat dual purpose in group theory. The order of a group $G$ is the number of elements it contains and and this value, also denoted $|G|$, need not be finite. On the other hand, we define the order of a group {\it element} $g$ to be the smallest $k>0$ such that $g^k=e$. If no such $k$ exists, then $g$ is said to have infinite order.

Probably the most familiar group is the set of all integers, denoted $\Z$, under the operation of addition. It has $1$ as its identity, inverses $-a$ for every whole number $a$, and the commutative property; likewise, any vector space $V$ is also a group under vector addition. The set of nonzero real numbers forms a group under multiplication. In these examples, the binary operation has the property that for any $g,h\in G$, the products $gh$ and $hg$ are equal. A group where this holds is called an {\it abelian} or {\it commutative} group. An example of a non-abelian group is the group $Q_8$ of quaternions with identity $1$, governed by the identities $(-1)^2 = 1$ and $i^2 = j^2 = k^2 = ijk = -1$. It has the following multiplication table:
% Right-aligned table with some extra spacing
$$\vcenter{\vbox{\offinterlineskip
    \halign{
        \hfil $#$ & \hfil\vrule# \hfil \enskip &
        \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ &
        \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ & \hfil $\;#\;$ \cr
        && 1 & -1 & i & -i & j & -j & k & - k \cr
        & height4pt &&&&&&& \cr
        \noalign{\hrule}
        & height4pt &&&&&&& \cr
        1 && 1 & -1 & i & -i & j & -j & k & - k \cr
        & height4pt &&&&&&& \cr
        -1 && -1 & 1 & -i & i & -j & j & -k & k \cr
        & height4pt &&&&&&& \cr
        i && i & -i & -1 & 1 & k & -k & -j & j \cr
        & height4pt &&&&&&& \cr
        -i && -i & i & 1 & -1 & -k & k & j & -j \cr
        & height4pt &&&&&&& \cr
        j && j & -j & -k & k & -1 & 1 & i & -i \cr
        & height4pt &&&&&&& \cr
        -j && -j & j & k & -k & 1 & -1 & -i & i \cr
        & height4pt &&&&&&& \cr
        k && k & -k & j & -j & -i & i & -1 & 1 \cr
        & height4pt &&&&&&& \cr
        -k && -k & k & -j & j & i & -i & 1 & -1 \cr
    }
}}$$
The general linear group $GL_n(\R)$ of invertible $n\times n$ matrices is another example of a non-abelian group, whenever $n>1$.

Group theory is intimately connected to the study of symmetries of an object. Let $T$ be a set and let $\Sym(T)$ denote the set of all bijections from $T$ to itself. This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $\Id_T$ and every bijection $f$ has an inverse bijection $f^{-1}$. In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections. For instance, $GL_n(\R) \subset \Sym(\R^n)$.

\section 2.2. Permutation Groups

A bijection $\sigma$ from a set $T$ to itself is also called a {\it permutation}. We will focus on the case where $T$ is finite and we may simply number the elements of $T=\{1,\ldots,n\}$. Then the set of permutations of $T$ is denoted $S_n$ and called the {\it symmetric group on $n$ letters} or, alternatively, the {\it permutation group on $n$ letters}. An element $\sigma$ of this group may be explicitly presented in such a way that we see where each element is taken to by $\sigma$:
$$\sigma = \pmatrix{1 & 2 & \cdots & n \cr \sigma(1) & \sigma(2) & \cdots & \sigma(n)}$$
It is easy to see that there are $n! = n(n-1)(n-2)\cdots 1$ different permutations of $n$ letters. There are $n$ choices for $\sigma(1)$; subsequently there remain $n-1$ choices for $\sigma(2)$, $n-2$ choices for $\sigma(3)$ and so on until our hand is forced for $\sigma(n)$. So $|S_n| = n!$.
Let us now consider the concrete example of the permutation
$$\sigma = \pmatrix{1 & 2 & 3 & 4 & 5 & 6 & 7 \cr 1 & 7 & 6 & 4 & 5 & 2 & 3},$$
which is an element of $S_7$. Notice that the elements 1, 4, and 5 are untouched by the permutation and the remaining four items are permuted in the cycle $2\mapsto 7$, $7\mapsto 3$, $3\mapsto 6$, and $6\mapsto 2$. This suggests a more concise notation for $\sigma$, since this cycle $(2\,7\,3\,6)$ completely determines the permutation. Notation-wise, $(2\,7\,3\,6)$ denotes exactly the same cycle as $(3\,6\,2\,7)$; to avoid confusion, we usually select the one that starts with the smallest number. Not every permutation is a cycle; for example the permutation
$$\tau = \pmatrix{1 & 2 & 3 & 4 \cr 4 & 3 & 2 & 1}$$
cannot be represented as a single cycle. However, it contains two cycles, $(1\,4)$ and $(2\,3)$, so we can simply represent $\tau$ as $(1\,4)(2\,3)$. When two cycles permute disjoint sets of elements, the cycles commute, so $(1\,4)(2\,3) = (2\,3)(1\,4)$. Every permutation can be represented as a product of disjoint cycles. In particular, the identity permutation can be represented as $(1)$, but we will often simply denote it $e$ or $()$.

\section 2.3. Subgroups

@@

We turn our attention to $S_3$. This group is not commutative, since the elements $\sigma = (1\,2\,3)$ and $\tau=(1\,2)$ do not commute. Multiplying $\sigma\tau$ we get $(1\,3)$ whereas the product $\tau\sigma = (2\,3)$.

Note that for $k\leq n$, $S_k\leq S_n$ because we can simply fix the letters $k+1,k+2,\ldots,n$. An easy corollary, then, is that $S_n$ does not commute for $n\geq 3$. This is because $S_3\leq S_n$ and we can simply take $\sigma$ and $\tau$ as elements of the larger group $S_n$ that do not commute.

@@

\section 2.4. Cosets and Normal Subgroups

Let $H$ be a subgroup of a group $G$. For any $g\in G$, we can form a {\it left coset} of $H$ by multiplying every element of $H$ by $g$ on the left:
$$gH = \{ gh : h\in H\}$$
Symmetrically, we could form a {\it right coset} $Hg$ of $H$. Now we prove that the cosets partition the group $G$.

\proclaim Lemma P. Let $H$ be a subgroup of a group $G$. Then every element of $G$ is in exactly one left coset of $H$.

\proof Let $g\in G$ be given. We commence by noting that since $H$ contains the identity element $e$, the element $g$ is in at least one coset, namely $gH$. Now we show that any two cosets are either disjoint or equal. Suppose $aH$ and $bH$ are two cosets that are not disjoint. This implies that there exist $h_1, h_2\in H$ such that $ah_1 = bh_2$. We can manipulate this identity to obtain $a^{-1}b = h_1h_2^{-1}$. So $a^{-1}b\in H$. Then from closure properties of subgroups, the cosets must be equal, since
$$aH = a(a^{-1}bH) = (aa^{-1})bH = bH.$$
Thus $g$ is in exactly one coset of $H$.\slug

Any two cosets have the same size, namely the size of $H$, since the for any coset $aH$, the function $f:H\rightarrow aH$ given by $a\mapsto aH$ establishes a bijection. This gives us some information about $H$, since it induces a partition of $G$ into equally-sized disjoint cosets. In fact, we have just proved the following famous theorem:

\thmparens L {Lagrange} Let $H$ be a subgroup of a group $G$. Then the order of $H$ must divide the order of $G$.\slug

Thus the value $|G|/|H|$ is an integer; it is called the {\it index} of $H$ in $G$ and denoted $[G:H]$. Lagrange's theorem also tells us that, for any element $g$ of a group $G$ with order $n$, the order of the cyclic subgroup $\langle g\rangle$ must divide $n$, so the order of the element $g$ divides $n$. Consequently, we know $g^n=e$ for any element $g\in G$.

We now know that a subgroup $H$ can partition a group $G$ in two ways, namely into left cosets of the form $aH$ or into right cosets of the form $Ha$. Remark that in general, the left cosets {\it do not equal} the right cosets.

\section 2.5. Isomorphisms and Homomorphisms

@@

Let $f:G\rightarrow G'$. The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e'$ of the target group $G'$. The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $G'$ that equal $f(g)$ for some $g\in G$. Then the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G'$. It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively. In fact, the kernel of a homomorphism is a normal subgroup.

@@

\section REFERENCES

The contents of this document are heavily based on the following three sets of lectures: Math 122 given by Benedict Gross at Harvard University, Fall 2003; MATH 235 given by Dani Wise at McGill University, Fall 2018; and MATH 456 given by Mika\"el Pichot at McGill University, Fall 2019.
